---
title: "ST237 Assignment 3"
author: "Eddie Heath"
date: "2025-03-15"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
#Reading libraries
#and running commands to install less common libraries, such as sf and geodata
knitr::opts_chunk$set(echo = TRUE)
install.packages("sf", repos = "https://cran.r-project.org")
install.packages("geodata", repos = "https://cran.r-project.org")
install.packages("readxl", repos = "https://cran.r-project.org")
install.packages("lubridate", repos = "https://cran.r-project.org")
install.packages("zoo", repos = "https://cran.r-project.org")
install.packages("scales", repos = "https://cran.r-project.org")
install.packages("tidyverse", repos = "https://cran.r-project.org")
install.packages("tidyr", repos = "https://cran.r-project.org")
install.packages("dplyr", repos = "https://cran.r-project.org")
library(readxl)
library(dplyr)
library(tidyverse)
library(tidyr)
library(lubridate)
library(zoo)
library(ggplot2)
library(sf)
library(scales)
library(geodata)
```

```{r 1A}
#Reading the dataset
CO2Data <- read_csv("ST237_co2_gdp_cleaned.csv",show_col_types=FALSE)

#Renaming the columns to useable names
colnames(CO2Data)[4] <- "Emissions"
colnames(CO2Data)[5] <- "GDP"
colnames(CO2Data)[6] <- "Population"
colnames(CO2Data)[7] <- "Region"

#Making a list of all the countries that have a label
Labels <- c("Singapore", "United Arab Emirates", "Saudi Arabia", "United States", 
           "Ireland", "South Korea", "United Kingdom", "Japan", "Russia", 
           "Malaysia", "Turkey", "China", "Belarus", "Mexico", "Costa Rica", 
           "Mongolia", "Indonesia", "Guatemala", "India", "Nigeria", 
           "Mozambique", "Ethiopia")

#Creating the plot
ggplot(data = CO2Data, aes(x = GDP, y = Emissions, size = Population, 
                           colour = Region)) +
  geom_point(alpha = 0.6) +
  #Setting up the x-axis
  scale_x_log10(breaks=c(2000,3000,4000,5000,6000,7000,8000,9000,10000,
                    20000,30000,40000,50000,60000,70000,80000,90000,100000),
                minor_breaks=c(1000,3000,4000,6000,7000,8000,9000,20000,30000,
                               40000,60000,70000,80000,90000),
                labels=c("$2,000","","","$5,000","","","","","$10,000",
                         "$20,000","","","$50,000","","","","","$100,000"))+
  #Setting up the y-axis
  scale_y_continuous(breaks=c(0,5,10,15,20,25,30),
    labels=c("0t","5t","10t","15t","20t","25t","30t"),
    minor_breaks = NULL)+
  #Creating the scale for the points being sized relative to population
  scale_size(range=c(1,12),
             guide=guide_legend(title="Circle sized by population"),
             breaks=c(600000000,1400000000),
             labels=c("600M","1.4B"))+
  #Adding text for the given countries, re-sizing for visibility
  geom_text(data = subset(CO2Data, Entity %in% Labels), 
            aes(label = Entity, size = (Population)/16+10000000), vjust=-1, 
            check_overlap = TRUE, show.legend = FALSE) +
  #Adding titles/captions etc
  labs(title="Consumption-based C02 emissions per capita vs. GDP per capita, 2022",
    subtitle="Consumption-based emissions ^1 are measured in tonnes per person. 
    They are territorial emissions minus emissions embedded in exports, 
    plus emissions embedded in imports.
GDP per capita is adjusted for price differences between countries (PPP) and 
over time (inflation).",
x="GDP per capita",
    y="Consumption-based emissions per capita",
captions="Data source: Global Carbon Budget (2024); Population based on various 
sources (2024); World Bank (2023)
OurWorldinData.org/co2-and-greenhouse-gas-emissions | CC BY

^1 Consumption-based emissions: Consumption-based emissions are national or 
regional emissions that have been adjusted for trade. They are 
calculated as domestic (or 'production-based' emissions) emissions minus the 
emissions generated in the production of goods and services that are 
exported to other countries or regions, plus emissions from the production of 
goods and services that are imported. Consumption-based emissions = 
Production-based - Exported + Imported emissions",
    color=NULL) +
  #Minimal theme to remove gridlines etc
  theme_minimal(base_size=15) +
  #Resizing various text elements
  theme(legend.position="right",
    plot.caption=element_text(size=6,hjust=0),
    plot.title = element_text(face="bold",size=10),
    plot.subtitle=element_text(size=6.5),
    legend.title=element_text(size=9),        
    legend.text=element_text(size=9),
    axis.title.x=element_text(size=10),
    axis.text.x=element_text(size=8),
    axis.line.x=element_line(color="black"),
    axis.title.y=element_text(size=10))
```

```{r 1B}
#Overall, I found the original plot quite good, and with relatively few issues 
#to improve. However, the two main problems I had were overcrowding, as there
#are too many data points to easily see all of them, and the skewed x-axis.
#In order to counter these issues, I chose to do the same plot as before, but
#faceted into 6 different graphs, one for each region. In order to preserve
#integrity, I made sure to keep the same scale for each region, such as to
#not be deceptive in presentation of the data.
#I also added more points for the legend of size relative to population,
#as the intial plot did not provide much useful info on this

ggplot(data = CO2Data, aes(x = GDP, y = Emissions, size = Population, 
                           colour = Region)) +
  geom_point(alpha = 0.6) +
  #Setting up the x-axis
  scale_x_continuous(breaks=c(0,10000,
                    20000,30000,40000,50000,60000,70000,80000,90000,100000,
                    110000,120000), 
                    labels = c("$0","$10,000","","","$40,000", "","","$70,000",
                               "", "", "$100,000","","$120,000"))+
  #Setting up the y-axis
  scale_y_continuous(breaks=c(0,5,10,15,20,25,30),
    labels=c("0t","5t","10t","15t","20t","25t","30t"),
    minor_breaks = NULL)+
  #Creating the scale for the points being sized relative to population
  scale_size(range=c(1,12),
             guide=guide_legend(title="Circle sized by population"),
             breaks=c(10000000, 50000000, 100000000, 
                      400000000, 600000000,1400000000),
             labels=c("10M","50M","100M","400M","600M","1.4B"))+
  #Creating 6 plots, one for each region
  facet_wrap(~ Region, ncol = 2) +
  #Themes, lines
  theme_minimal() +
  theme(panel.grid.major = element_line(colour = "grey"),
    axis.text.x = element_text(size = 8, angle = 45, hjust=1))+ 
  guides(color = "none", size = guide_legend(title = "Population")) +
  #Labels
  labs(title = "Consumption-based C02 emissions per capita vs. GDP per capita, 
       2022, separated by region", x = "GDP per capita", 
       y = "Consumption based emissions per capita")
```

\begin{center}
\small Data source: Global Carbon Budget (2024); Population based on various sources (2024); World Bank (2023)
OurWorldinData.org/co2-and-greenhouse-gas-emissions | CC BY

Consumption-based emissions: Consumption-based emissions are national or regional emissions that have been adjusted for trade. They are 
calculated as domestic (or 'production-based' emissions) emissions minus the emissions generated in the production of goods and services that are 
exported to other countries or regions, plus emissions from the production of goods and services that are imported. Consumption-based emissions = 
Production-based - Exported + Imported emissions
\end{center}

```{r 2A}
#Reading Data
covidData <- read_xlsx("region_newDeaths28DaysByDeathDate.xlsx")

#Removing unnecessary columns, creating a rolling average column
covidData <- covidData %>%
  dplyr::select(-c(metric, metric_name, area_type, area_code)) %>%
  mutate(date = as.Date(date)) %>%
  group_by(area_name) %>%
  complete(date = seq.Date(min(date, na.rm = TRUE), max(date, na.rm = TRUE), 
                           by = "day")) %>%
  ungroup() %>%
  mutate(value = replace_na(value, 0))%>%
  group_by(area_name) %>%
  mutate(WeekAvg = rollapply(value, width = 7, FUN = mean, fill = NA, 
                             align = "right", na.rm = TRUE)) %>%
  ungroup()%>%
  mutate(WeekAvg = replace_na(WeekAvg, 0))

#Renaming columns for ease of use
colnames(covidData)[1] <- "Region"

#Grouping data by date to create totals
covidDataTotal <- covidData%>%
  group_by(date) %>%
  summarise(Total = sum(value, na.rm = TRUE))

#Setting up custom colours for nicer viewing
regionColours <- c("East Midlands" = "#a6cee3", "East of England" = "#1f78b4", 
                   "South East" = "#b2df8a", "London" = "#fb9a99", 
                   "North West" = "#33a02c", "South West" = "#e31a1c", 
                   "Yorkshire and The Humber" = "#fdbf6f", "North East" = "#ff7f00",
                   "West Midlands" = "#cab2d6")

#When using geom_area for this graph, it creates a graph with a zero-line
#and all data above this line, whereas the original plot has the data
#'floating' with no line. Whilst my method may not be true to this, the only
#method I could find to create data like this was the geom_stream library,
#which cannot handle the number of data points contained in the data set, and I
#chose to prioritize keeping all data values over plotting the data in this manner

#Creating the plot
ggplot(covidData, aes(x = date, y = WeekAvg, fill = Region)) +
  geom_area() +
  #Adding in only breaks for the year
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  #Titles, captions, etc
  labs(title = "More than 1,000 deaths each day are attributed to Covid-19",
    subtitle = "Daily deaths attributed to Covid-19 (7-day rolling average)",
    x = NULL, y = NULL,
    caption = "Source: COVID-19 Archive data download, supplied by the 
    UK Health Security Agency (UKHSA)") +
  theme_minimal() +
  #Adding custom colours
  scale_fill_manual(values = regionColours) +
  #Adjusting text size, etc
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(),
        plot.title = element_text(size = 16),
        plot.subtitle = element_text(size = 12),
        plot.caption = element_text(size = 7),
        panel.grid = element_blank(),
        legend.position = "bottom")+
  #Adding lines for chosen key dates
  #Start of lockdown
  geom_vline(xintercept = as.Date("2020-03-03"), linetype = "dashed") +
  geom_text(aes(x = as.Date("2020-03-03"), y = -30,
                label = "A"), size = 2.5, 
            angle = 90, vjust = 1.2) +
  #Vaccine Release
  geom_vline(xintercept = as.Date("2020-12-08"), linetype = "dashed") +
  geom_text(aes(x = as.Date("2020-12-08"), y = -30,
                label = "B"), size = 2.5, 
            angle = 90, vjust = 1.2) +
  #Peak Deaths
  geom_vline(xintercept = as.Date("2021-01-19"), linetype = "dashed") +
  geom_text(aes(x = as.Date("2021-01-19"), y = -30,
                label = "C"), size = 2.5, 
            angle = 90, vjust = 1.2) +
  #End of Lockdowns
  geom_vline(xintercept = as.Date("2021-02-21"), linetype = "dashed") +
  geom_text(aes(x = as.Date("2021-02-21"), y = -30,
                label = "D"), size = 2.5, 
            angle = 90, vjust = 1.2) +
  #Covid declassified
  geom_vline(xintercept = as.Date("2023-05-05"), linetype = "dashed") +
  geom_text(aes(x = as.Date("2023-05-05"), y = -30,
                label = "E"), size = 2.5, 
            angle = 90, vjust = 1.2)
```

\begin{center}
\small A = Initial Lockdown 3rd March 2020
B = Initial Vaccine 8th December 2020
C = Peak Deaths 19th January 2021
D = End of Restrictions 21st February 2021
E = Official End of Pandemic 5th May 2023
\end{center}

```{r 2BSetup, message=FALSE, warning=FALSE}
#Reading population data
popData <- read_xlsx("regionalpopestimatesenglandandwales19712022.xlsx", 
                     sheet = "Table 4")

#Renaming columns for ease of use
colnames(popData)[3] <- "Region"
colnames(popData)[4] <- "Population"
popData$Region[popData$Region == "East"] <- "Eastern"

#Filtering out only English data for 2022
popData <- popData %>% filter(Contents == "2022")
popData <- popData %>% filter(Region != "Wales")
popData <- popData[c(3,4)]

#Reading data for a map
ukRegionMapAvg <- st_read("https://martinjc.github.io/UK-GeoJSON/json/eng/topo_eer.json", 
                          quiet = TRUE)
ukRegionMapCapita <- st_read("https://martinjc.github.io/UK-GeoJSON/json/eng/topo_eer.json", 
                             quiet = TRUE)

#Adding a tag for periods of 6 months
covidData <- covidData %>%
  mutate(date = as.Date(date),
         # Create a period label; here we assume you have 6 periods
         Time = cut(date, breaks = "6 months", labels = paste("Period", 1:7)))

#Total deaths for a region
totalDeaths <- covidData %>%
  group_by(Region) %>%
  summarise(Total = sum(value, na.rm = TRUE))

#Averaging over 6 months
sixMonthsAvg <- covidData %>%
  group_by(Time, Region) %>%
  summarise(Total = mean(value, na.rm = TRUE))

#And again for another set of plots
sixMonthsCapita <- covidData %>%
  group_by(Time, Region) %>%
  summarise(Total = mean(value, na.rm = TRUE))

#Again renaming columns
colnames(totalDeaths)[1] <- "Region"
colnames(ukRegionMapAvg)[4] <- "Region"
colnames(ukRegionMapCapita)[4] <- "Region"
totalDeaths$Region[totalDeaths$Region == "East of England"] <- "Eastern"

#Making the data set wide for combining
sixMonthsAvg <- sixMonthsAvg %>%
  pivot_wider(names_from = Time, values_from = Total)

#And again for the separate dataset
sixMonthsCapita <- sixMonthsCapita %>%
  pivot_wider(names_from = Time, values_from = Total)

#Renaming Regions to match
sixMonthsAvg$Region[sixMonthsAvg$Region == "East of England"] <- "Eastern"
sixMonthsCapita$Region[sixMonthsCapita$Region == "East of England"] <- "Eastern"

#Combining with population data to calculate per capita deaths
sixMonthsCapita <- sixMonthsCapita %>%
  left_join(popData, by = "Region") %>%
  pivot_longer(cols = starts_with("Period"), names_to = "Time", values_to = "Value")

#Calculating per capita deaths
sixMonthsCapita$Value <- as.numeric(sixMonthsCapita$Value)
sixMonthsCapita$Population <- as.numeric(sixMonthsCapita$Population)
sixMonthsCapita$Capita <- sixMonthsCapita$Value/sixMonthsCapita$Population

#Making the data set wide again for combining
sixMonthsCapita <- sixMonthsCapita %>%
  pivot_wider(names_from = Time, values_from = c(Value, Population, Capita))

#Removing unnecessary columns
sixMonthsCapita <- sixMonthsCapita[c(1,16,17,18,19,20,21,22)]

#Combining map with covid data
ukRegionMapAvg <- ukRegionMapAvg %>%
  left_join(sixMonthsAvg, by = "Region")

#And making the dataset long for plotting
ukRegionMapAvg <- ukRegionMapAvg %>%
  pivot_longer(cols = starts_with("Period"), names_to = "Time", values_to = "Value")

#And again for per capita data
ukRegionMapCapita <- ukRegionMapCapita %>%
  left_join(sixMonthsCapita, by = "Region")

ukRegionMapCapita <- ukRegionMapCapita %>%
  pivot_longer(cols = starts_with("Capita_Period"), names_to = "Capita", values_to = "Value")

ukRegionMapCapita$Value <- ukRegionMapCapita$Value*1000000
```

\newpage

```{r 2BGraph1, fig.height=8, fig.width=6}
#Creating labels for the different data periods
labelsAvg = c("Period 1" = "Mar - Aug 2020", "Period 2" = "Sep2020 - Feb2021", 
              "Period 3" = "Mar - Aug 2021", "Period 4" = "Sep2021 - Feb2022",
              "Period 5" = "Mar - Aug 2022", "Period 6" = "Sep2022 - Feb2023",
              "Period 7" = "Mar - Aug 2023")

#Creating the plots
ggplot(data = ukRegionMapAvg) +
  geom_sf(aes(fill = Value), color = "black") +
  theme_minimal() +
  #Creating a facet wrap that produces 7 related plots in 1
  facet_wrap(~ Time, ncol = 3, labeller = labeller(Time = labelsAvg)) +
  #Renaming Legend
  scale_fill_gradient(low = "white", high = "red", name = "No. of Deaths per Day",
                      breaks = c(0,20,40,60), limits = c(0,70)) +
  #Title, caption, etc
  labs(title = "Average No. of Covid Deaths per Day over 6-month Periods", 
  caption = "Graphs showing the progression of total deaths due to covid
       over time, with data averaged over a 6 month period") +
  #Adjusting text
  theme(axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(),
        axis.ticks.x = element_blank(), 
        axis.text.x = element_blank(),
        strip.text = element_text(size = 7),
        legend.position = "bottom")
```

\begin{center}
\small Source: COVID-19 Archive data download, supplied by the 
    UK Health Security Agency (UKHSA)
    
\small March - Aug 2020 containing initial lockdown
    Sep2020 - Feb2021 containing initial vaccine, peak deaths, end of restrictions
    March - Aug 2023 containing end of pandemic
\end{center}

```{r 2BGraph2, fig.height=8, fig.width=6}
#Same as above but for per capita dataset

ggplot(data = ukRegionMapCapita) +
  geom_sf(aes(fill = Value), color = "black") +
  theme_minimal() +
  facet_wrap(~ Capita, ncol = 3, labeller = labeller(Time = labelsAvg)) +
  scale_fill_gradient(low = "white", high = "red",
                      name = "No. of Deaths per Million People per Day") +
  labs(title = "Avg No. of Covid Deaths per Million People per Day over 6-month Periods") +
  theme(axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(),
        axis.ticks.x = element_blank(), 
        axis.text.x = element_blank(),
        strip.text = element_text(size = 7),
        legend.position = "bottom")+
  labs(caption = "Graphs showing the progression of average deaths per millions
  people due to covid over time, with data averaged over a 6 month period and 
  then divided by the total population for the area (Using 2022 population data)")
```
\begin{center}
\small Source: COVID-19 Archive data download, supplied by the 
    UK Health Security Agency (UKHSA)
    
 \small  March - Aug 2020 containing initial lockdown
    Sep2020 - Feb2021 containing initial vaccine, peak deaths, end of restrictions
    March - Aug 2023 containing end of pandemic
\end{center}






